---
title: "Simulate many data sets"
author: "Ada Yan"
date: "`r Sys.Date()`"
output: html_document
---

To do: rerun the inference with the new simulated data
```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = FALSE, fig.width = 8, fig.height = 8)
```

```{r include = FALSE}
rm(list = ls())
git_dir <- "../"
devtools::load_all(git_dir)
library(tidyverse)
library(cmdstanr) # switch to cmdstanr for now as require Stan 2.34 for dirichlet_multinomial which is not available on rstan
library(bayesplot)
set.seed(1)
dir_name <- make_dir_Rmd("simulate_data.Rmd")
samples_filename <- paste0(dir_name, "samples_immune.rds")
```

# Aims

- Use LHS sampling to find a range of parameter values to produce reasonable viral load trajectories
- Also, for the same parameter set, we want to generate different noisy data sets (see. Fig. 5 of https://doi.org/10.1016/j.epidem.2017.01.004 for visualisation of many replicates)
- To be comparable between data sets, we fix innate and adaptive immune parameters to single values

```{r pars, echo = TRUE}
pars <- get_default_pars()

pars[["burst_size"]] <- pars[["p"]] / pars[["delta"]]

log10_pars <- log10(pars)
names(log10_pars) <- paste0("log10_", names(pars))

varied_par_names <- c("log10_beta1", "log10_burst_size", "log10_V_0", "log10_delta", "log10_c1", "alpha")

# 1 order of magnitude each way for everything except $\alpha$
varied_par_ranges <- matrix(1, ncol = 2, nrow = length(varied_par_names))
varied_par_ranges[,1] <- -1
varied_par_ranges <- varied_par_ranges + log10_pars[varied_par_names]
varied_par_ranges[nrow(varied_par_ranges),] <- c(0, 1)
rownames(varied_par_ranges) <- varied_par_names
colnames(varied_par_ranges) <- c("min", "max")
N_samples <- 5000
N_samples_target <- 100
```

```{r eval = FALSE}
samples <- lhs::randomLHS(N_samples, length(varied_par_names))
colnames(samples) <- varied_par_names

scale_samples <- function(vec, bounds) {
  vec * diff(bounds) + bounds[1]
}

for(i in seq_along(varied_par_names)) {
  samples[,i] <- scale_samples(samples[,i], varied_par_ranges[i,])
}

log10_par_idx <- grepl("log10_", varied_par_names)
varied_par_names <- gsub("log10_", "", varied_par_names)

samples[,log10_par_idx] <- 10^samples[,log10_par_idx]
colnames(samples) <- varied_par_names

calc_r_wrapper <- function(beta1, T_0, p, delta, c1) {
  pars <- list(beta1 = beta1, T_0 = T_0, p = p, delta = delta, c1 = c1)
  calc_r(pars)
}
calc_r_wrapper <- Vectorize(calc_r_wrapper)

samples <- samples %>%
  as_tibble() %>%
  mutate(p = delta * burst_size,
         T_0 = 7e7,
         R_0 = beta1 * T_0 * p / delta / (c1 + pars[["k1"]] * pars[["A_0"]]), # for the homogeneous model
         r = calc_r_wrapper(beta1, T_0, p, delta, c1 + pars[["k1"]] * pars[["A_0"]])) %>% # for the homogeneous model
  select(-burst_size) %>%
  filter(R_0 > 1, r < log(1e5), r > log(1e3)) %>%
  slice_head(n = N_samples_target) %>%
  mutate(idx = seq_len(n()))

if(file.exists(samples_filename)) warning("overwriting existing parameter values")
saveRDS(samples, samples_filename)
```

# Simulations from model

```{r}
samples <- readRDS(samples_filename)
solving_time <- get_solving_time()

model_filename <- "innate_adaptive.R"
# compile ODEs
gen <- compile_model(model_filename)

varied_par_names <- gsub("log10_", "", varied_par_names)
fixed_par_names <- names(pars)[!(names(pars) %in% varied_par_names)]
fixed_par_names <- fixed_par_names[!fixed_par_names == "p"]

solve_model <- function(idx) {
  samples %>%
    filter(idx == !!idx) %>%
    select(-c(T_0, R_0, r, idx, alpha)) %>%
    as.list() %>%
    c(as.list(pars[fixed_par_names])) %>%
    simple_solve(gen = gen, solving_time = solving_time) %>%
    mutate(idx = !!idx)
} 
```

```{r}
sol <- lapply(seq_len(N_samples_target), solve_model) %>%
  bind_rows()

plot_sol <- function(start_idx) {
  sol %>%
    filter(idx >= start_idx, idx < start_idx + 20) %>%
    ggplot(aes(x = t, y = V)) +
    facet_wrap(~idx) +
    geom_line() +
    scale_y_log10("Viral load (pfu/mL)", limits = c(1, 1e10)) +
    theme_bw() +
    xlab("Time (days)")
}
```

```{r eval = FALSE, sol1, fig.cap = "Solutions to simulations 1-20."}
plot_sol(1)
```

```{r eval = FALSE, sol2, fig.cap = "Solutions to simulations 21-40."}
plot_sol(21)
```

```{r eval = FALSE, sol3, fig.cap = "Solutions to simulations 41-60."}
plot_sol(41)
```

```{r eval = FALSE, sol4, fig.cap = "Solutions to simulations 61-80."}
plot_sol(61)
```

```{r eval = FALSE, sol5, fig.cap = "Solutions to simulations 81-100."}
plot_sol(81)
```

# With heterogeneity

Note that in the WIMSIG simulations, I was varying $\beta_i$ and $\alpha = T_{01} / T_0$ independently, whereas now I am trying to keep $\bar{\beta} T_0 = \sum_{i=1}^N \beta_i T_{0i}$ constant.

First attempt at definition:

We define $\beta_1 = \bar{\beta} (1 - \Delta T_0/T_{01})$, $\beta_2 = \bar{\beta} (1 + \Delta T_0/T_{02})$ such that $\beta_1 T_{01} + \beta_2 T_{02} = (\bar{\beta} - \Delta \bar{\beta} T_{0}/T_{01}) T_{01} + (\bar{\beta} + \Delta \bar{\beta} T_{0}/T_{02}) T_{02} = \bar{\beta} T_0$ is constant regardless of the value of $\Delta$.

The range of $\Delta$ such that $\beta_1 > 0, \beta_2 > 0$ is $-T_{02}/T_0 < \Delta < T_{01}/T_0$.  We define a scaled version of $\Delta$, $\Delta'$, where $-1 \leq \Delta' \leq 1$. The scaling is defined by $\Delta = \Delta' T_{02}/T_0$ for $\Delta' < 0$ and $\Delta = \Delta' T_{01}/T_0$ for $\Delta' > 0$.  This leads to the desired ranges $-T_{02}/T_0 < \Delta < T_{01}/T_0$.  If we substitute the definition of $\Delta'$ into the definitions of $\beta_1$ and $\beta_2$, we get $\beta_1 = \bar{\beta} (1 - \Delta' T_{02}/T_{01})$, $\beta_2 = \bar{\beta} (1 + \Delta')$ for $\Delta' < 0$, $\beta_1 = \bar{\beta} (1 - \Delta'T_{01}/T_{02})$, $\beta_2 = \bar{\beta} (1 + \Delta')$ for $\Delta' > 0$.

In summary:

We define $\beta_1 = \bar{\beta} (1 - \Delta' T_{02}/T_{01})$, $\beta_2 = \bar{\beta} (1 + \Delta')$ for $\Delta' < 0$, $\beta_1 = \bar{\beta} (1 - \Delta')$, $\beta_2 = \bar{\beta} (1 + \Delta' T_{01}/T_{02})$ for $\Delta' > 0$, $-1 \leq \Delta' \leq 1$.

Note that $\Delta'$ is not directly the standard deviation for $\beta_i T_{0i}$, as $\beta_1 T_{01} = \bar{\beta} T_{01} - \bar{\beta} \Delta T_{0}$ and $\beta_2 T_{02} = \bar{\beta} T_{02} + \bar{\beta} \Delta T_{0}$. The standard deviation is actually $\bar{\beta} \frac{T_{02} - T_{01}}{2} + \bar{\beta} \Delta T_0$.

```{r}
Delta_scaled <- seq(-.8, .8, by = .2)

two_beta_filename <- "innate_adaptive_two_beta.R"
# compile ODEs
gen <- compile_model(two_beta_filename)

solve_two_beta <- function(idx, Delta_scaled) { # calculate beta1 T_01 + beta2 T_02 = beta T_0
  sample_pars <- samples %>%
    filter(idx == !!idx) %>%
    select(-c(T_0, R_0, r, idx)) %>%
    as.list()
  if(Delta_scaled < 0) {
    Delta <- Delta_scaled * (1 - sample_pars$alpha)
  } else {
    Delta <- Delta_scaled * sample_pars$alpha
  }
  beta1 <- sample_pars$beta1
  sample_pars$beta1[1] <-  beta1 * (1 - Delta / sample_pars$alpha)
  sample_pars$beta1[2] <-  beta1 * (1 + Delta / (1 - sample_pars$alpha))
  stopifnot(all(sample_pars$beta1 > 0))
  sample_pars <- c(sample_pars, as.list(pars[fixed_par_names]))
  simple_solve(gen = gen, pars = sample_pars, solving_time = solving_time) %>%
    mutate(idx = !!idx,
           Delta_scaled = !!Delta_scaled)
} 
```

```{r}
par_grid <- expand_grid(idx = seq_len(N_samples_target), Delta_scaled = Delta_scaled)
sol <- par_grid %>%
  apply_named_args(1, solve_two_beta) %>%
  bind_rows()

plot_subset <- c(-.8, 0, .8)

plot_sol <- function(start_idx, n_sims = 20) {
  g <- sol %>%
    filter(idx >= start_idx, idx < start_idx + n_sims, Delta_scaled %in% plot_subset) %>%
    ggplot(aes(x = t, y = V, color = Delta_scaled, group = Delta_scaled)) +
    geom_line() +
    scale_y_log10("Viral load (pfu/mL)", limits = c(1, 1e10)) +
    scale_colour_gradient2(low = "red",  mid = "black",  high = "blue") +
    theme_bw() +
    xlab("Time (days)") +
    labs(colour = latex2exp::TeX("$\\Delta$"))
  if(n_sims <= 5) {
    g <- g + facet_wrap(~idx, nrow = 1)
  } else {
    g <- g + facet_wrap(~idx)
  }
  g
}
```

```{r sol1-het, fig.cap = "Solutions to simulations 1-20."}
plot_sol(1)
```

```{r sol2-het, fig.cap = "Solutions to simulations 21-40."}
plot_sol(21)
```

```{r sol2-het-25, fig.cap = "Solutions to simulations 21-25.", fig.width = 8, fig.height = 2}
plot_sol(21, 5)
```

```{r sol3-het, fig.cap = "Solutions to simulations 41-60."}
plot_sol(41)
```

```{r sol4-het, fig.cap = "Solutions to simulations 61-80."}
plot_sol(61)
```

```{r sol5-het, fig.cap = "Solutions to simulations 81-100."}
plot_sol(81)
```

```{r}
plot_sol_cell <- function(start_idx) {
  sol %>%
    filter(idx >= start_idx, idx < start_idx + 5, Delta_scaled %in% plot_subset) %>%
    select(-V) %>%
    mutate(across(ends_with("]"), ~ .x / (`T1[1]` + `T1[2]` + `I1[1]` + `I1[2]`))) %>%
    pivot_longer(ends_with("]"), names_to = "compartment_name", values_to = "prop_value") %>%
    ggplot(aes(x = t, y = prop_value, colour = Delta_scaled, group = Delta_scaled)) +
    facet_grid(idx~compartment_name) +
    geom_line() +
    theme_bw() +
    xlab("Time (days)") +
    scale_y_log10("Proportion of cells") +
    scale_colour_gradient2(low = "red",  mid = "black",  high = "blue") +
    coord_cartesian(xlim = c(0, 2), ylim = c(1e-5, 1)) +
    labs(colour = latex2exp::TeX("$\\Delta$"))
}
```

```{r eval = FALSE, cell1, fig.cap = "Cell proportion data for simulations 1-5."}
plot_sol_cell(1)
```

```{r eval = FALSE, cell2, fig.cap = "Cell proportion data for simulations 6-10."}
plot_sol_cell(6)
```

```{r eval = FALSE, cell3, fig.cap = "Cell proportion data for simulations 11-15."}
plot_sol_cell(11)
```

```{r eval = FALSE, cell4, fig.cap = "Cell proportion data for simulations 16-20."}
plot_sol_cell(16)
```

```{r eval = FALSE, cell5, fig.cap = "Cell proportion data for simulations 21-25."}
plot_sol_cell(21)
```

```{r eval = FALSE, cell6, fig.cap = "Cell proportion data for simulations 26-30."}
plot_sol_cell(26)
```

```{r eval = FALSE, cell7, fig.cap = "Cell proportion data for simulations 31-35."}
plot_sol_cell(31)
```

```{r eval = FALSE, cell8, fig.cap = "Cell proportion data for simulations 36-40."}
plot_sol_cell(36)
```

```{r eval = FALSE, cell9, fig.cap = "Cell proportion data for simulations 41-45."}
plot_sol_cell(41)
```

```{r eval = FALSE, cell10, fig.cap = "Cell proportion data for simulations 46-50."}
plot_sol_cell(46)
```

```{r eval = FALSE, cell11, fig.cap = "Cell proportion data for simulations 51-55."}
plot_sol_cell(51)
```

```{r eval = FALSE, cell12, fig.cap = "Cell proportion data for simulations 56-60."}
plot_sol_cell(56)
```

```{r eval = FALSE, cell13, fig.cap = "Cell proportion data for simulations 61-65."}
plot_sol_cell(61)
```

```{r eval = FALSE, cell14, fig.cap = "Cell proportion data for simulations 66-70."}
plot_sol_cell(66)
```

```{r eval = FALSE, cell15, fig.cap = "Cell proportion data for simulations 71-75."}
plot_sol_cell(71)
```

```{r eval = FALSE, cell16, fig.cap = "Cell proportion data for simulations 76-80."}
plot_sol_cell(76)
```

```{r eval = FALSE, cell17, fig.cap = "Cell proportion data for simulations 81-85."}
plot_sol_cell(81)
```

```{r eval = FALSE, cell18, fig.cap = "Cell proportion data for simulations 86-90."}
plot_sol_cell(86)
```

```{r eval = FALSE, cell19, fig.cap = "Cell proportion data for simulations 91-95."}
plot_sol_cell(91)
```

```{r eval = FALSE, cell20, fig.cap = "Cell proportion data for simulations 96-100."}
plot_sol_cell(96)
```

Put noise on the simulated data:

```{r}
sampling_times <- get_sampling_times()
sampling_pars <- get_sampling_pars()
```

```{r}
simulate_data_partial <- function(idx, Delta_scaled, eval) {
  sim_filename <- paste0(dir_name, "sim_data_immune_", idx, "_",
                         make_filename_from_number(Delta_scaled), ".rds")
  if(!eval) return(sim_filename)
  sol %>%
    filter(idx == !!idx, Delta_scaled == !!Delta_scaled) %>%
    simulate_data(sampling_times, sampling_pars, cells = TRUE, sim_filename)
}
```

```{r eval = FALSE}
par_grid %>%
  mutate(eval = TRUE) %>%
  apply_named_args(1, simulate_data_partial)
```

```{r}
get_sim_data <- function(idx, Delta_scaled) {
  sim_data <- simulate_data_partial(idx, Delta_scaled, eval = FALSE) %>% 
    readRDS()
  sim_data$obs_V <- sim_data$obs_V %>%
    mutate(idx = !!idx, Delta_scaled = !!Delta_scaled)
  sim_data$obs_C <- sim_data$obs_C %>%
    mutate(idx = !!idx, Delta_scaled = !!Delta_scaled)
  sim_data
}

sim_data <- apply_named_args(par_grid, 1, get_sim_data)
sim_data_V <- lapply(sim_data, \(x) x$obs_V) %>%
  bind_rows()
sim_data_C <- lapply(sim_data, \(x) x$obs_C) %>%
  bind_rows()
rm(sim_data)

plot_sol_V_noise <- function(sol, sim_data_V, start_idx) {
  sol <- sol %>%
    filter(idx >= start_idx, idx < start_idx + 20, Delta_scaled %in% plot_subset)
  sim_data_V <- sim_data_V %>%
    filter(idx >= start_idx, idx < start_idx + 20, Delta_scaled %in% plot_subset)
  ggplot(sol, aes(x = t, color = Delta_scaled, group = Delta_scaled)) +
    facet_wrap(~idx) +
    geom_line(aes(y = V)) +
    geom_point(data = sim_data_V, aes(y = obs_V)) +
    scale_y_log10("Viral load (pfu/mL)", limits = c(1, 1e10)) +
    scale_colour_gradient2(low = "red",  mid = "black",  high = "blue") +
    theme_bw() +
    xlab("Time (days)") +
    labs(colour = latex2exp::TeX("$\\Delta$"))
}
```


```{r fig.width = 4, fig.height = 3, eval = FALSE}
plot_sol_V_noise(sol, sim_data_V, 100) 
```

```{r sol-V-noise1, fig.cap = "Solutions to viral load simulations 1-20 with noise."}
plot_sol_V_noise(sol, sim_data_V, 1) 
```

```{r sol-V-noise2, fig.cap = "Solutions to viral load simulations 21-40 with noise."}
plot_sol_V_noise(sol, sim_data_V, 21) 
```

```{r sol-V-noise3, fig.cap = "Solutions to viral load simulations 41-60 with noise."}
plot_sol_V_noise(sol, sim_data_V, 41) 
```

```{r sol-V-noise4, fig.cap = "Solutions to viral load simulations 61-80 with noise."}
plot_sol_V_noise(sol, sim_data_V, 61) 
```

```{r sol-V-noise5, fig.cap = "Solutions to viral load simulations 81-100 with noise."}
plot_sol_V_noise(sol, sim_data_V, 81) 
```

```{r}
plot_sol_cell <- function(start_idx) {
  
  sim_data_C <- sim_data_C %>%
    rename(`U1[1]` = `T1[1]`,
           `U1[2]` = `T1[2]`)
  keep_cols <- colnames(sim_data_C)
  
  process <- function(sol) {
    sol <- sol %>%
      filter(idx >= start_idx, idx < start_idx + 5, Delta_scaled %in% plot_subset)#
    if("R[1]" %in% colnames(sol)) {
      sol <- sol %>%
        mutate(`U1[1]` = `T1[1]` + `R[1]`,
               `U1[2]` = `T1[2]` + `R[2]`)
      
    }
    
    sol %>%
      select(all_of(keep_cols)) %>%
      mutate(across(ends_with("]"), ~ .x / (`U1[1]` + `U1[2]` + `I1[1]` + `I1[2]`))) %>%
      pivot_longer(ends_with("]"), names_to = "compartment_name", values_to = "prop_value") %>%
      filter(Delta_scaled %in% plot_subset)
  }
  
  sol <- sol %>% process()
  sim_data_C <- sim_data_C %>% process()
  
  ggplot(sol, aes(x = t, y = prop_value, colour = Delta_scaled, group = Delta_scaled)) +
    facet_grid(idx~compartment_name) +
    geom_line() +
    geom_point(data = sim_data_C) +
    theme_bw() +
    xlab("Time (days)") +
    scale_y_log10("Proportion of cells") +
    scale_colour_gradient2(low = "red",  mid = "black",  high = "blue") +
    coord_cartesian(xlim = c(0, 2), ylim = c(1e-5, 1)) +
    labs(colour = latex2exp::TeX("$\\Delta$"))
}
```

```{r cell-noise1, fig.cap = "Cell proportion data for simulations 1-5."}
plot_sol_cell(1)
```

```{r cell-noise2, fig.cap = "Cell proportion data for simulations 6-10."}
plot_sol_cell(6)
```

```{r cell-noise3, fig.cap = "Cell proportion data for simulations 11-15."}
plot_sol_cell(11)
```

```{r cell-noise4, fig.cap = "Cell proportion data for simulations 16-20."}
plot_sol_cell(16)
```

```{r cell-noise5, fig.cap = "Cell proportion data for simulations 21-25."}
plot_sol_cell(21)
```

```{r cell-noise6, fig.cap = "Cell proportion data for simulations 26-30."}
plot_sol_cell(26)
```

```{r cell-noise7, fig.cap = "Cell proportion data for simulations 31-35."}
plot_sol_cell(31)
```

```{r cell-noise8, fig.cap = "Cell proportion data for simulations 36-40."}
plot_sol_cell(36)
```

```{r cell-noise9, fig.cap = "Cell proportion data for simulations 41-45."}
plot_sol_cell(41)
```

```{r cell-noise10, fig.cap = "Cell proportion data for simulations 46-50."}
plot_sol_cell(46)
```

```{r cell-noise11, fig.cap = "Cell proportion data for simulations 51-55."}
plot_sol_cell(51)
```

```{r cell-noise12, fig.cap = "Cell proportion data for simulations 56-60."}
plot_sol_cell(56)
```

```{r cell-noise13, fig.cap = "Cell proportion data for simulations 61-65."}
plot_sol_cell(61)
```

```{r cell-noise14, fig.cap = "Cell proportion data for simulations 66-70."}
plot_sol_cell(66)
```

```{r cell-noise15, fig.cap = "Cell proportion data for simulations 71-75."}
plot_sol_cell(71)
```

```{r cell-noise16, fig.cap = "Cell proportion data for simulations 76-80."}
plot_sol_cell(76)
```

```{r cell-noise17, fig.cap = "Cell proportion data for simulations 81-85."}
plot_sol_cell(81)
```

```{r cell-noise18, fig.cap = "Cell proportion data for simulations 86-90."}
plot_sol_cell(86)
```

```{r cell-noise19, fig.cap = "Cell proportion data for simulations 91-95."}
plot_sol_cell(91)
```

```{r cell-noise20, fig.cap = "Cell proportion data for simulations 96-100."}
plot_sol_cell(96)
```

```{r fig.height = 2.5, fig.width = 8, fig.cap = "Cell proportion data for simulations 96-100."}
plot_sol_cell(100)
```
```{r}
```

