---
title: "Fit heterogeneous model, including cell type data, with LHS"
author: "Ada Yan"
date: "`r Sys.Date()`"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE)
```

```{r include = FALSE}
rm(list = ls())
git_dir <- "../"
devtools::load_all(git_dir)
library(cmdstanr) # switch to cmdstanr for now as require Stan 2.34 for dirichlet_multinomial which is not available on rstan
library(bayesplot)
library(tidyverse)
sim_dir_name <- make_dir_Rmd("simulate_data.Rmd")
dir_name <- make_dir_Rmd("05_fit_cell_data_lhs.Rmd")
set.seed(1)
```

# Use simulated data from simulate_data.Rmd

```{r}
default_pars <- get_default_pars()
samples <- readRDS(paste0(sim_dir_name, "samples_immune.rds"))
sampling_pars <- get_sampling_pars()
```

```{r}
varied_par_names <- c("log10_beta1", "log10_burst_size", "log10_V_0", "log10_delta", "log10_c1", "alpha")
varied_par_names <- gsub("log10_", "", varied_par_names)
fixed_par_names <- names(default_pars)[!(names(default_pars) %in% varied_par_names)]
fixed_par_names <- fixed_par_names[!fixed_par_names == "p"]
```

# Fitting

## Run Stan

```{r}
Delta_scaled <- c(-.8, 0, .8)

simulate_data_partial <- function(idx, Delta_scaled, eval) {
  sim_filename <- paste0(sim_dir_name, "sim_data_immune_", idx, "_",
                         make_filename_from_number(Delta_scaled), ".rds")
  if(!eval) return(sim_filename)
  sol %>%
    filter(idx == !!idx, Delta_scaled == !!Delta_scaled) %>%
    simulate_data(sampling_times, sampling_pars, cells = TRUE, sim_filename)
}

fit_wrapper <- function(idx, Delta_scaled, inference_model, adapt_delta = NULL, eval = TRUE, chains = 4) {
  sim_filename <- simulate_data_partial(idx, Delta_scaled, eval = FALSE)
  
  if(is.null(adapt_delta)) {
    fit_filename <- gsub("sim_data", "fit3", sim_filename)
  } else {
    fit_filename <- gsub("sim_data", "refit3", sim_filename)
  }
  fit_filename <- gsub(sim_dir_name, dir_name, fit_filename)
  
  if(!eval) return(fit_filename)
  
  sim_data <- sim_filename %>% readRDS()
  
  sim_data$obs_C <- sim_data$obs_C %>%
    filter(t == 2)
  # times at which data is observed, plus 0
  solving_times <- unique(c(sim_data$obs_V$t, sim_data$obs_C$t)) %>% c(0) %>% unique() %>% sort()
  # for stan solver exclude 0
  solving_times_stan <- setdiff(solving_times, 0)
  
  match_idx <- function(obs_times, solving_times) {
    match(obs_times, solving_times) - 1
  }
  
  data_list <- samples %>%
    filter(idx == !!idx) %>%
    select(V_0, delta, c1, p, alpha) %>%
    as.list() %>%
    c(list(sigma_V = sampling_pars$sigma_V,
           lod_V = sampling_pars$lod_V,
           rho0 = sampling_pars$rho,
           T_V = nrow(sim_data$obs_V),
           T_C = nrow(sim_data$obs_C),
           T_S = length(solving_times_stan),
           ts = solving_times_stan,
           t_V_idx = match_idx(sim_data$obs_V$t, solving_times),
           t_C_idx = match_idx(sim_data$obs_C$t, solving_times),
           obs_V = sim_data$obs_V$obs_V,
           obs_C = t(as.matrix(select(sim_data$obs_C, ends_with("]")))),
           evaluate_likelihood = 1),
      as.list(default_pars[fixed_par_names]))
  
  fit <- inference_model$sample(
    data = data_list,
    seed = 1,
    chains = chains,
    parallel_chains = chains,
    adapt_delta = adapt_delta
  )
  fit$save_object(file = fit_filename)
  invisible(NULL)
}

par_grid <- expand_grid(idx = seq_len(nrow(samples)), Delta_scaled = Delta_scaled)

filename_fn <- \(idx, Delta_scaled)  fit_wrapper(idx, Delta_scaled, eval = FALSE)
```

```{r eval = FALSE}
inference_model <- cmdstan_model(get_model_filename("two_beta_innate_adaptive_cells"))
fit_partial <- function(idx, Delta_scaled) fit_wrapper(idx, Delta_scaled, inference_model, adapt_delta = NULL, eval = TRUE, chains = 8)
apply_named_args(par_grid, 1, fit_partial)
```

## Exclude chains that didn't converge or had divergent transitions

```{r eval = FALSE}
get_summaries_subset_chains_partial <- \(idx, Delta_scaled) get_summaries_subset_chains(idx, Delta_scaled, fit_wrapper)
summaries <- apply_named_args(par_grid, 1, get_summaries_subset_chains_partial) %>%
  bind_rows()
saveRDS(summaries, paste0(dir_name, "summaries_refit3.rds"))

```


# Inferred Delta

```{r eval = FALSE}
inferred_Delta_scaled <- apply_named_args(par_grid, 1, \(idx, Delta_scaled) get_pars(filename_fn = filename_fn,
                                                                                     par_names = "Delta_scaled", idx = idx, Delta_scaled = Delta_scaled)) %>%
  bind_rows() %>%
  mutate(chain = as.numeric(chain))

# exclude the same chains as above
summaries_refit_temp <- summaries %>%
  select(idx, Delta_scaled, chosen_chain_combn) %>%
  rename(chain = chosen_chain_combn)

inferred_Delta_scaled <- inferred_Delta_scaled %>%
  inner_join(summaries_refit_temp)
saveRDS(inferred_Delta_scaled, paste0(dir_name, "inferred_Delta_scaled3.rds"))
```

```{r inferred-Delta-scaled, fig.cap = "Inferred value of $\\Delta\'$ (mean and 95% CI) vs. true value of $\\Delta\'$ (top of facet, red line), by true value of $\\beta$ (x-axis).", fig.width = 6, fig.height = 2.5}
inferred_Delta_scaled <- readRDS(paste0(dir_name, "inferred_Delta_scaled3.rds"))

quantiles_Delta_scaled <- inferred_Delta_scaled %>%
  select(-iter, -chain) %>%
  group_by(idx, Delta_scaled) %>%
  summarise(lower = quantile(value, probs = .025, na.rm = TRUE),
            median = quantile(value, probs = .5, na.rm = TRUE),
            upper = quantile(value, probs = .975, na.rm = TRUE))

appender <- function(string) 
  latex2exp::TeX(paste("$\\Delta = $", string))  

quantiles_Delta_scaled %>%
  full_join(select(samples, idx, beta1)) %>%
  drop_na() %>%
  mutate(Delta_scaled = round(Delta_scaled, 1)) %>%
  ggplot() +
  facet_wrap(~Delta_scaled, nrow = 1, labeller = as_labeller(appender, 
                                                             default = label_parsed)) +
  geom_point(aes(x = beta1, y = median)) +
  geom_hline(aes(yintercept = Delta_scaled), color = "red") +
  geom_errorbar(aes(x = beta1, ymin = lower, ymax = upper)) +
  theme_bw() +
  scale_x_log10(latex2exp::TeX("True value of \\beta"), guide = guide_axis(angle = 90)) +
  ylab(latex2exp::TeX("Inferred value of $\\Delta$")) +
  theme(panel.spacing = unit(1, "lines"))
```


Number of simulations where 95% CIs excluded 0:

```{r}
quantiles_Delta_scaled %>%
  mutate(exclude_0 = upper < 0 | lower > 0) %>%
  group_by(Delta_scaled) %>%
  summarise(prop_exclude_0 = mean(exclude_0))
```

Visually see that no CIs excluding 0 were in the wrong direction

# Model predictions to show why identifiability is needed

Make model predictions varying $\alpha$ from the value for each simulation. First do this for the true values

```{r}
solving_time <- get_solving_time()

two_beta_filename <- "innate_adaptive_two_beta.R"
# compile ODEs
gen <- compile_model(two_beta_filename)
```

```{r}
solve_two_beta <- function(idx, Delta_scaled, alpha) { # calculate beta1 T_01 + beta2 T_02 = beta T_0
  # the beta need to be calculated with the original alpha
  sample_pars <- samples %>%
    filter(idx == !!idx) %>%
    select(-c(T_0, R_0, r, idx)) %>%
    as.list()
  if(Delta_scaled < 0) {
    Delta <- Delta_scaled * (1 - sample_pars$alpha)
  } else {
    Delta <- Delta_scaled * sample_pars$alpha
  }
  beta1 <- sample_pars$beta1
  sample_pars$beta1[1] <-  beta1 * (1 - Delta / sample_pars$alpha)
  sample_pars$beta1[2] <-  beta1 * (1 + Delta / (1 - sample_pars$alpha))
  stopifnot(all(sample_pars$beta1 > 0))
  # then change to the new alpha
  sample_pars$alpha <- alpha
  sample_pars <- c(sample_pars, as.list(default_pars[fixed_par_names]))
  simple_solve(gen = gen, pars = sample_pars, solving_time = solving_time) %>%
    mutate(idx = !!idx,
           Delta_scaled = !!Delta_scaled,
           alpha = !!alpha)
}

alpha_vec <- seq(0, 1, by = .1)

```

Now make the model predictions (95\% CI) for the model fits:

```{r}
make_extended_obs_data_list <- function(data_list) {
  data_list$ts <- seq(0, 7, by = .01) %>% setdiff(0)
  data_list$T_S <- length(data_list$ts)
  data_list
}

make_pred_new_alpha <- function(idx, Delta_scaled, alpha, eval) {
  fit_filename <- filename_fn(idx, Delta_scaled)
  fit <- readRDS(fit_filename)
  pred_filename <- gsub("/fit", paste0("/pred_", make_filename_from_number(alpha)), fit_filename)
  if(!eval) return(pred_filename)
  
  data_list <- samples %>%
    filter(idx == !!idx) %>%
    select(V_0, delta, c1, alpha, p) %>%
    as.list()
  
  # get betas from the big tibble that has all the results
  N_samples <- 100
  betas <- betas %>%
    filter(idx == !!idx, Delta_scaled == !!Delta_scaled) %>%
    pivot_wider(names_from = par_name, values_from = value) %>%
    select(starts_with("beta")) %>%
    slice_sample(n = N_samples)
  
  data_list <- make_extended_obs_data_list(data_list)
  data_list <- c(data_list, as.list(default_pars[fixed_par_names]))
  data_list$beta1 <- as.matrix(betas)
  data_list$N_samples <- N_samples
  data_list$alpha <- alpha
  
  pred_model <- cmdstan_model(get_model_filename("two_beta_innate_adaptive_betas_pred"))
  
  pred <- pred_model$sample(
    data = data_list,
    seed = 1,
    chains = 1,
    fixed_param = TRUE,
    iter_warmup = 1,
    iter_sampling = 1
  )
  # browser()
  pred$save_object(file = pred_filename)
  invisible(NULL)
}
```

```{r eval = FALSE}
par_grid <- expand_grid(par_grid, alpha = alpha_vec)
apply_named_args(par_grid, 1, \(idx, Delta_scaled, alpha) make_pred_new_alpha(idx, Delta_scaled, alpha, eval = TRUE))
```

Plot them against the true values:

```{r fig.width = 8, fig.height = 3}
plot_preds_different_Delta_scaled_alpha_true <- function(idx, Delta_scaled_vec, alpha_vec, facets = FALSE) {

  par_grid <- expand_grid(Delta_scaled = Delta_scaled_vec,
                          alpha = alpha_vec)
  
  sols <- par_grid %>%
    apply_named_args(1, \(Delta_scaled, alpha) solve_two_beta(idx, Delta_scaled, alpha)) %>%
    bind_rows() %>%
    mutate(alpha = factor(alpha))
  
  T_0_lab <- "Initial proportion of \n cell type 1"
  
  appender <- function(string) 
    latex2exp::TeX(paste("$\\Delta = $", string))  
  
  original_alpha <- samples %>%
    filter(idx == !!idx) %>%
    pull(alpha) %>%
    round(digits = 2)
  
  g <- ggplot(sols) +
    # geom_ribbon(aes(x = t, ymin = lower, ymax = upper, fill = alpha, group = alpha), alpha = .5) +
    geom_line(aes(x = t, y = V, color = alpha, group = alpha)) +
    scale_y_log10(latex2exp::TeX("Viral load (TCID$_{50}$)"), limits = c(1, 1e8), breaks = 10^seq(0, 8, by = 2)) +
    theme_bw() +
    xlab("Time (days)") +
    labs(color = latex2exp::TeX(T_0_lab))#
  
  if(facets) {
    g <- g + facet_grid(alpha~Delta_scaled)
  } else {
    g <- g + facet_wrap(~Delta_scaled, nrow = 1, labeller = as_labeller(appender, 
                                                                        default = label_parsed))
  }
  g
}
```

```{r eval = TRUE, fig.width = 8, fig.height = 3}
plot_preds_different_Delta_scaled_alpha_true(1, Delta_scaled, seq(.2, .8, by = .2))
```


```{r fig.width = 8, fig.height = 3}
plot_preds_different_Delta_scaled_alpha <- function(idx, Delta_scaled_vec, alpha_vec, facets = FALSE) {
  get_pred_different_Delta_scaled_alpha <- function(Delta_scaled, alpha) {
    data_list <- make_extended_obs_data_list(list())
    pred_times <- data_list$ts
    # par_grid_subset <- par_grid %>%
    #   filter(idx %in% remaining_large_error,
    #          Delta_scaled %in% Delta_scaled_subset)
    pred_filename <- make_pred_new_alpha(idx, Delta_scaled, alpha, eval = FALSE)
    sol_V <- get_sol_V(pred_filename, pred_times, used_pred = TRUE)
    quantiles <- sol_V %>%
      group_by(t) %>%
      summarise(lower = quantile(V, prob = 0.025),
                median = quantile(V, prob = 0.5),
                upper = quantile(V, prob = 0.975)) %>%
      ungroup() %>%
      mutate(Delta_scaled = !!Delta_scaled,
             alpha = !!alpha)
    quantiles
  }
  
  par_grid <- expand_grid(Delta_scaled = Delta_scaled_vec,
                          alpha = alpha_vec)
  
  quantiles <- par_grid %>%
    apply_named_args(1, get_pred_different_Delta_scaled_alpha) %>%
    bind_rows() %>%
    mutate(alpha = factor(alpha))
  
  # sols <- par_grid %>%
  #   apply_named_args(1, \(Delta_scaled, alpha) solve_two_beta(idx, Delta_scaled, alpha)) %>%
  #   bind_rows() %>%
  #   mutate(alpha = factor(alpha))
  
  T_0_lab <- "Initial proportion of \n cell type 1"
  
  appender <- function(string) 
    latex2exp::TeX(paste("$\\Delta = $", string))  
  
  original_alpha <- samples %>%
    filter(idx == !!idx) %>%
    pull(alpha) %>%
    round(digits = 2)
  
  exclude_0 <- quantiles_Delta_scaled %>%
    filter(idx == !!idx) %>%
    mutate(exclude_0 = upper < 0 | lower > 0) %>%
    pull(exclude_0)
  
  my_title <- paste(c(idx, original_alpha, exclude_0), collapse = " ")
  
  g <- ggplot(quantiles) +
    
    geom_ribbon(aes(x = t, ymin = lower, ymax = upper, fill = alpha, group = alpha), alpha = .5) +
    # geom_line(aes(x = t, y = V, color = alpha, group = alpha)) +
    scale_y_log10(latex2exp::TeX("Viral load (TCID$_{50}$)"), limits = c(1, 1e8), breaks = 10^seq(0, 8, by = 2)) +
    theme_bw() +
    xlab("Time (days)") +
    labs(color = latex2exp::TeX(T_0_lab), fill = latex2exp::TeX(T_0_lab))# +
  ggtitle(my_title)
  
  if(facets) {
    g <- g + facet_grid(alpha~Delta_scaled)
  } else {
    g <- g + facet_wrap(~Delta_scaled, nrow = 1, labeller = as_labeller(appender, 
                                                                        default = label_parsed))
  }
  g
}
```

```{r eval = TRUE, fig.width = 8, fig.height = 3}
plot_preds_different_Delta_scaled_alpha(1, Delta_scaled, seq(.2, .8, by = .2))
```

Compare individual simulations:

```{r fig.height = 8, fig.width = 8}
plot_preds_different_Delta_scaled_alpha_individual <- function(idx, Delta_scaled_vec, alpha_vec, facets = FALSE) {
  n_sims_to_plot <- 5
  sims_to_plot <- sample.int(100, size = n_sims_to_plot)
  
  get_pred_different_Delta_scaled_alpha <- function(Delta_scaled, alpha) {
    data_list <- make_extended_obs_data_list(list())
    pred_times <- data_list$ts
    # par_grid_subset <- par_grid %>%
    #   filter(idx %in% remaining_large_error,
    #          Delta_scaled %in% Delta_scaled_subset)
    pred_filename <- make_pred_new_alpha(idx, Delta_scaled, alpha, eval = FALSE)
    sol_V <- get_sol_V(pred_filename, pred_times, used_pred = TRUE) %>%
      filter(idx %in% sims_to_plot) %>%
      mutate(Delta_scaled = !!Delta_scaled,
             alpha = !!alpha)
    sol_V
  }
  
  par_grid <- expand_grid(Delta_scaled = Delta_scaled_vec,
                          alpha = alpha_vec)
  sol_pred <- par_grid %>%
    apply_named_args(1, get_pred_different_Delta_scaled_alpha) %>%
    bind_rows() %>%
    mutate(alpha = factor(alpha),
           Delta_scaled = paste("$\\Delta = $", Delta_scaled),
           idx = paste("Sample $", idx, "$"))
  
  sols <- par_grid %>%
    apply_named_args(1, \(Delta_scaled, alpha) solve_two_beta(idx, Delta_scaled, alpha)) %>%
    bind_rows() %>%
    mutate(Delta_scaled = paste("$\\Delta = $", Delta_scaled),
           alpha = factor(alpha),
           idx = "$\\textrm{True}$ $\\textrm{value}$") %>%
    bind_rows(sol_pred)
  
  T_0_lab <- "Initial proportion of\\ cell type 1"
  
  appender <- function(string)  latex2exp::TeX(string)
  
  original_alpha <- samples %>%
    filter(idx == !!idx) %>%
    pull(alpha) %>%
    round(digits = 2)
  
  exclude_0 <- quantiles_Delta_scaled %>%
    filter(idx == !!idx) %>%
    mutate(exclude_0 = upper < 0 | lower > 0) %>%
    pull(exclude_0)
  
  # my_title <- paste(c(idx, original_alpha, exclude_0), collapse = " ")
  my_title <- paste("Simulation", idx)
  
  g <- ggplot(sols) +
    facet_grid(idx~Delta_scaled, labeller = as_labeller(appender, 
                                                        default = label_parsed)) +
    geom_line(aes(x = t, y = V, color = alpha, group = alpha)) +
    scale_y_log10(latex2exp::TeX("Viral load (TCID$_{50}$)"), limits = c(1, 1e8), breaks = 10^seq(0, 8, by = 2)) +
    theme_bw() +
    xlab("Time (days)") +
    labs(color = latex2exp::TeX(T_0_lab), fill = latex2exp::TeX(T_0_lab)) +
    ggtitle(my_title)
  
  g
}
```

```{r eval = TRUE, fig.width = 8, fig.height=8}
g <- lapply(seq_len(10), \(idx) plot_preds_different_Delta_scaled_alpha_individual(idx, Delta_scaled, seq(.2, .8, by = .2)))
g
```


```{r}
sessionInfo()
```

This report is brought to you by `fluheterogeneousmodels` commit `r system("git -C ~/git_repos/fluheterogeneousmodels/ rev-parse --short HEAD", intern = TRUE)`.
